---
title: "college_final_project"
format: html
execute:
  http: true
---

```{r}
# install.packages("httr2")
library(httr2)
library(tidyverse)
library(plotly)
```

```{r}
school_list <- list()

for (k in 1:500) {
college_request <- request("https://api.data.gov") %>% 
  req_url_path_append("ed") %>% 
  req_url_path_append("collegescorecard") %>% 
  req_url_path_append("v1") %>% 
  req_url_path_append("schools") %>% 
  req_url_query(
    api_key = "sXz5twfrusSSiWtD5EQ35d4e3XNzfiw3m4yHiMuv",
    fields = "id,school.name,latest.student.size,school.state,latest.cost.tuition.in_state,latest.cost.tuition.out_of_state,latest.admissions.admission_rate.overall,latest.cost.attendance.academic_year,latest.completion.completion_rate_4yr_150nt,latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid,latest.earnings.1_yr_after_completion.median,latest.earnings.5_yrs_after_completion.median" ,
    per_page = 100,
    page = k,
  )


  college_response <- req_perform(college_request)
  data <- college_response %>% resp_body_json(simplifyVector = TRUE)
  results <- data$results
  school_list[[k]] <- results
  
  if (nrow(results) < 100) {
    break
  }
}
# install.packages("janitor")

# college_tbl <- college_response %>%
  # resp_body_json(simplifyVector = TRUE) %>%
  # pluck("results") %>%
  # as_tibble()

# college_tbl

colleges <- bind_rows(school_list)

```

```{r}
# colnames(colleges)
```

```{r}

# not sure how accurate the apps & accept columns
# could use this set too it has more rows:
# https://www.kaggle.com/datasets/thedevastator/u-s-department-of-education-college-scorecard-da
#college2 <- read_csv("College_Data.csv")
#college2

#college2 <- college2 %>%
# rename(name = ...1)

#colleges <- colleges %>%
#  rename(name = school.name)
```

```{r}
# merged_colleges <- left_join(college2, colleges, by = "name")

# some duplicates in here
```

```{r}
alt_colleges <- read_csv("alt_college_data.csv")
```

```{r}
alt_colleges <- alt_colleges %>%
  rename(name = INSTNM)

alt_colleges <- alt_colleges %>%
  rename(id = UNITID)
```

```{r}
# merged_colleges_big <- left_join(colleges, alt_colleges, by = c("school.name" = "name"))
# merged_colleges_big <- left_join(colleges, alt_colleges, by = "id")
```

```{r}
# merged_colleges %>%
#  ggplot(aes(x = Private, y = latest.cost.tuition.out_of_state)) +
#    geom_boxplot() +
#    coord_flip() +
#    theme_minimal() +
#    labs(
#      y = "Tuition (Out of State)",
#      x = "Private",
#      title = "Cost of Public vs. Private Schools"
#    )
```

# Part 3: ROI

```{r}
colleges <- colleges %>%
  rename(name = school.name) %>%
  rename(state = school.state) %>%
  rename(five_yr_earnings = latest.earnings.5_yrs_after_completion.median) %>%
  rename(one_yr_earnings = latest.earnings.1_yr_after_completion.median) %>%
  rename(size = latest.student.size) %>%
  rename(instate_tuition = latest.cost.tuition.in_state) %>%
  rename(outstate_tuition = latest.cost.tuition.out_of_state) %>%
  rename(acceptance = latest.admissions.admission_rate.overall) %>%
  rename(avg_cost = latest.cost.attendance.academic_year) %>%
  rename(graduation = latest.completion.completion_rate_4yr_150nt) %>%
  rename(repaid_debt = latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid)
```

## Variables in question

`name`: the institution name
`state`: the institution's state
`instate_tuition`: cost of instate tuition, before aid
`outstate_tuition`: cost of out-of-state tuition, before aid
`acceptance`: overall acceptance rate
`graduation`: overall graduation rate, in four years
`repaid_debt`: proportion of graduates who have fully repaid debt after 2 years
`size:` the most recent number of undergraduate students
`avg_cost:` average cost of attendance, before aid
`one_yr_earnings:` median earnings of graduates after 1 year from graduation
`five_yr_earnings:` median earnings of graduates after 5 years from graduation

## Charts

```{r}
roi_plot1 <- colleges %>%
  ggplot(aes(x = avg_cost, y = graduation, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Graduation Rate",
      color = "Acceptance Rate",
      title = "Graduation Rates vs. Cost of Attendance"
    )
ggplotly(roi_plot1, tooltip = "text")
```

### Observation: there is a clear positive trend between the two variables. This suggests that more expensive colleges might be worth it because there is a higher chance of graduating with a degree.

```{r}
roi_plot2 <- colleges %>%
  ggplot(aes(x = avg_cost, y = one_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 1 Year",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 1 Year"
    )
ggplotly(roi_plot2, tooltip = "text")
```

### Observation: there is a slight positive trend, but it is almost flat. This suggests that a more expensive college might not be worth it in the short-term, using 1-year median earnings as a marker of ROI.

```{r}
roi_plot3 <- colleges %>%
  ggplot(aes(x = avg_cost, y = five_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 5 Years",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 5 Years"
    )
ggplotly(roi_plot3, tooltip = "text")
```

### Observation: the trend is more significant after 5 years. This gives evidence that a more expensive college might be worth it in the long-term, as they can lead to higher-paying careers. 

```{r}
roi_plot4 <- colleges %>%
  ggplot(aes(x = avg_cost, y = acceptance)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Acceptance Rate",
      title = "Acceptance Rate vs. Cost of Attendance",
    )
roi_plot4
```






### Observation: there is a strong negative association between these variables. More expensive colleges are often more selective, with lower acceptance rates. This could mean that the cost of attendance doesn't necessarily predict outcomes, but the prestige of the institution.




#Part 4

#What effect does income status have on the benefits recieved from a university/major?

```{r}

library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)

#Variables in question:
#Major/Field of study = Program title
#Median earnings of pell grant recipients 4 years after graduating = earnings_4_yr_median_pell
#Median earnings of non pell grant recipients 4 years after graduating = earnings_4_yr_median_nonpell



url ="https://api.data.gov/ed/collegescorecard/v1/schools?api_key=sydxfG3MlGadt1EOtAGWOSuSMnYBVFFLvkBeAFiE&school.state=CA&latest.admissions.admission_rate.overall__range=..0.40&latest.programs.cip_4_digit.credential.level=3&fields=id,school.name,school.enrollment.all,latest.programs.cip_4_digit.code,latest.programs.cip_4_digit.title,latest.programs.cip_4_digit.credential.level,latest.programs.cip_4_digit.earnings.4_yr.overall_median_earnings,latest.programs.cip_4_digit.earnings.4_yr.pell_median_earnings,latest.programs.cip_4_digit.earnings.4_yr.nonpell_median_earnings,latest.programs.cip_4_digit.earnings.5_yr.working_not_enrolled.overall_count,latest.cost.attendance.academic_year,latest.programs.cip_4_digit.earnings.4_yr.working_not_enrolled.pell_count,latest.programs.cip_4_digit.earnings.4_yr.working_not_enrolled.nonpell_count,latest.student.size&page=0&per_page=100"

response <- GET(url)

data <- fromJSON(content(response, "text", encoding = "UTF-8"), flatten = TRUE)

ca_schools2 <- data$results
ca_flat2 <- ca_schools2 %>%
  unnest(latest.programs.cip_4_digit)


ca_clean2 <- ca_flat2 %>%
  rename(
    school_id = id,
    school_name = school.name,
    cip_code = code,
    program_title = title,
    population = latest.student.size,
    degree_level = credential.level,
    earnings_4_yr_median = earnings.4_yr.overall_median_earnings,
    earnings_4_yr_median_pell = earnings.4_yr.pell_median_earnings,
    earnings_4_yr_median_nonpell = earnings.4_yr.nonpell_median_earnings,
    count_5_yr_overall = earnings.5_yr.working_not_enrolled.overall_count, 
    count_4_yr_pell = earnings.4_yr.working_not_enrolled.pell_count,       
    count_4_yr_nonpell = earnings.4_yr.working_not_enrolled.nonpell_count,
    tuiton = latest.cost.attendance.academic_year
  )
  




```

```{r}
majors<- ca_clean2%>%
group_by(cip_code,program_title)%>%
summarize(
total_population = sum(count_5_yr_overall, na.rm = TRUE),
majorearningspell = mean(earnings_4_yr_median_pell,na.rm = TRUE), majorearningsnonpell =mean(earnings_4_yr_median_nonpell,na.rm = TRUE),majorearningsoverall =mean(earnings_4_yr_median,na.rm = TRUE),.groups="drop")%>%
mutate(privelege_tax = (majorearningsnonpell - majorearningspell)
  ,privelege_percentage = (privelege_tax/majorearningsoverall))
  top_10_tax <-majors %>%
  filter(total_population>100)%>%
  slice_max(order_by = privelege_tax, n = 10)
  not_10_tax <-majors %>%
    filter(total_population>100)%>%
  slice_min(order_by = privelege_tax, n = 10)
ggplot(top_10_tax, aes(x = reorder(program_title, privelege_tax), y = privelege_tax)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Majors by Earnings difference between Pell and non Pell recipients.",
    subtitle = "Amount less earned by Pell Grant Recipients",
    x = "Major",
    y = "Earnings Gap ($)"
  )
ggplot(not_10_tax, aes(x = reorder(program_title, privelege_tax), y = privelege_tax)) +
  geom_col(fill = "firebrick") + 
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Majors by Earnings difference between Pell and non Pell recipients.",
    subtitle = "Amount less earned by Pell Grant Recipients",
    x = "Major",
    y = "Earnings Gap ($)"
  )

ggplot(top_10_tax, aes(x = reorder(program_title, privelege_percentage), y = privelege_percentage*100)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Majors by least earnings difference between Non Pell and Pell students",
    subtitle = "Percent less earned by Pell Grant Recipients",
    x = "Major",
    y = "Earnings Gap (%)"
  )





```



#Observation:The gaps in career earnings are very significant but its hard to detect a major trend in the types of majors with the highest gaps.I expected more meritocratic "skills based" majors to have lower gaps, while more relationships oriented majors would feature large spreads, however there is truly a variety of majors represented. Engineering in particular is featured in multiple spots in both the highest earnings gap and lowest earnings gap majors.



```{r}

popular_majors <- ca_clean2 %>%
  filter(!is.na(count_5_yr_overall), population > 0,tuiton>0) %>%
  group_by(school_name) %>%
  slice_max(order_by = count_5_yr_overall, n = 5, with_ties = FALSE) %>%
  mutate(privelege_tax = earnings_4_yr_median_nonpell - earnings_4_yr_median_pell,privelege_tax_pct = (privelege_tax / earnings_4_yr_median_nonpell) * 100)

school_summaries <- popular_majors %>%
  summarize(
    tuiton = first(tuiton),
    avg_privelege_tax = mean(privelege_tax, na.rm = TRUE),
    avg_privelege_tax_pct = mean(privelege_tax_pct,na.rm=TRUE),
    .groups = "drop"
  )

top_10_schools <- school_summaries %>%
  slice_max(order_by = avg_privelege_tax, n = 10)

ggplot(top_10_schools, aes(x = reorder(school_name, avg_privelege_tax), y = avg_privelege_tax)) +
  geom_col(fill = "firebrick") + 
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Selective California Schools Pell Earning Gaps",
    subtitle = "Average difference in Non-Pell vs Pell earnings across the 5 most popular majors",
    x = "School Name",
    y = "Average Earnings Gap ($)"
  )


ggplot(school_summaries, aes(x = tuiton, y = avg_privelege_tax_pct)) +
  geom_point(color = "firebrick", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, color = "black", se = TRUE, fill = "gray80") +
  theme_minimal() +
  labs(
    title = "Tuition Cost vs. Pell Earnings Gap",
    subtitle = " regression of total academic year cost against average percent difference between Pell earnings",
    x = "Cost of Attendance ($)",
    y = "Average Pell Earnings gap Gap ($)"
  )
```
#Observation: no clear correlation between the Pell-Grant earnings gap percentage(among the schools most popular majors) and Attendance cost, although very constrained by lack of institutional level data. The school level data is also hard to tell much from given the amount of variance, a big difficulty is the people who benefit the most from their backrounds will not file for finacial aid and thus will be excluded from our dataset. Also notably this dataset is comprised of only selective California Universities, which due to the incredibly strong public university options in the state could lead to a more evenly distributed wealth of students across schools of different tuition prices then in other states.

```{r}
url ="https://api.data.gov/ed/collegescorecard/v1/schools?api_key=sydxfG3MlGadt1EOtAGWOSuSMnYBVFFLvkBeAFiE&&latest.admissions.admission_rate.overall__range=..0.30&latest.programs.cip_4_digit.credential.level=3&fields=id,school.name,school.enrollment.all,latest.programs.cip_4_digit.code,latest.programs.cip_4_digit.title,latest.programs.cip_4_digit.credential.level,latest.programs.cip_4_digit.earnings.1_yr.overall_median_earnings,latest.programs.cip_4_digit.earnings.4_yr.pell_median_earnings,latest.programs.cip_4_digit.earnings.4_yr.nonpell_median_earnings,latest.programs.cip_4_digit.earnings.5_yr.working_not_enrolled.overall_count,latest.cost.attendance.academic_year,latest.programs.cip_4_digit.earnings.4_yr.working_not_enrolled.pell_count,latest.programs.cip_4_digit.earnings.4_yr.working_not_enrolled.nonpell_count,latest.programs.cip_4_digit.earnings.highest.1_yr.working_not_enrolled.overall_count,latest.student.size,latest.programs.earnings.1_yr.nonpell_median_earnings,latest.programs.cip_4_digit.earnings.1_yr.pell_median_earnings,latest.programs.cip_4_digit.earnings.1_yr.nonpell_median_earnings&page=0&per_page=100"

response <- GET(url)

data <- fromJSON(content(response, "text", encoding = "UTF-8"), flatten = TRUE)

biz_schools2 <- data$results
biz <- biz_schools2 %>%
  unnest(latest.programs.cip_4_digit)


biz_clean <- biz %>%
  rename(
    school_id = id,
    school_name = school.name,
    cip_code = code,
    program_title = title,
    population = latest.student.size,
    degree_level = credential.level,
    earnings_1_yr_median = earnings.1_yr.overall_median_earnings,
    earnings_4_yr_median_pell = earnings.4_yr.pell_median_earnings,
    earnings_1_yr_median_pell = earnings.1_yr.pell_median_earnings,
    earnings_4_yr_median_nonpell = earnings.4_yr.nonpell_median_earnings,
    earnings_1_yr_median_nonpell = earnings.1_yr.nonpell_median_earnings,
    count_5_yr_overall = earnings.5_yr.working_not_enrolled.overall_count, 
    count_4_yr_pell = earnings.4_yr.working_not_enrolled.pell_count,       
    count_4_yr_nonpell = earnings.4_yr.working_not_enrolled.nonpell_count,
    count_1_year=earnings.highest.1_yr.working_not_enrolled.overall_count,
    tuiton = latest.cost.attendance.academic_year
  )
biz_clean<- biz_clean%>%
filter(as.numeric(cip_code) >= 5202 & as.numeric(cip_code) <= 5215)%>%
group_by(school_name)


nonscorecard<- university_salary_data%>%
mutate(Buisnessesschool = str_replace(Institution, "\\s*\\(.*", ""))
bizcomparison <- inner_join(biz_clean, nonscorecard, by = c("school_name" = "Buisnessesschool"), suffix = c("Govdata","Bizscholdata"))%>%
 filter(!is.na(earnings_1_yr_median))%>%
group_by(school_name) %>%
  mutate(
    totalpop = sum(count_1_year, na.rm = TRUE),
    earnings_1_yr_average = sum((count_1_year* earnings_1_yr_median) / sum(count_1_year, na.rm = TRUE), na.rm = TRUE)
  ) %>% 
  ungroup()%>%
  mutate(
difference =(base_average_salary-earnings_1_yr_average),na.rm=TRUE)

databiz<-bizcomparison%>%
distinct(school_name, .keep_all = TRUE)%>%
mutate(bizschool = mean(base_average_salary))%>%
mutate(grinder= mean(earnings_1_yr_average))%>%
mutate(meandiff = mean(difference))


```



```{r}
datatwiz<-databiz%>%
distinct(grinder,bizschool)
p <-ggplot(databiz, aes(x=reorder(school_name,difference), y =difference,text=paste0(
        "Tuition:  ", tuiton,
        " <br> Earnings Difference<br>", difference,"<br> First Year Median Finacial Aid Recipient,Pell Grant Recipient earnings::",earnings_1_yr_median_pell,"<br> First Year Median Finacial Aid Recipient, Non-Pell Grant Recipient earnings:",earnings_4_yr_median_nonpell)))+
  geom_col(fill = "firebrick") + 
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Biz School Data to the Rescue?",
    subtitle = "Difference in undergraduate Buisness School 1st year salary reports and College Scorecard 1st year Median Earnings",
    x = "School Name",
    y = "Average Earnings Gap ($)"
  )+
  theme(legend.position ="right")
ggplotly(p, tooltip = "text")
ggplot(datatwiz)+
  geom_col(aes(x = "Biz School Mean", y = bizschool), fill = "firebrick", width = 0.4) + 
  geom_col(aes(x = "Gov Scorecard Mean", y = grinder), fill = "steelblue", width = 0.4) + 
  theme_minimal() +
  labs(
    title = "Biz School Data to the Rescue?",
    subtitle = "Same school Difference in undergraduate Buisness School 1st year salary reports College Scorecard report",
    x = "Mean Earnings of Finacial Aid Recipient vs Overall School Average",
    y = "Average Earnings($)"
  )+
scale_y_continuous(
    labels = scales::dollar_format(),
    limits = c(0, 100000),
    breaks = seq(0, 100000, by = 50000))
```
# Observation: As you can the mean/median earnings of all graduates from this subset of Buisness schools who reported salary statistics(some variance in which statistic was reported) is significantly higher then the median from college scorecard among students in the same programs at the same schools.I think this data is helpful in providing some access into the world of non finacial aid student outcomes,as the college scorecard data forces comparisons between the best and worst off of students who recieve some finacial aid, when in reality the question of if the gains of expensive colleges accrue more towards wealthy students is based on students who despite also not recieving a pellgrant, have much higher levels of family household income then anyone in the college scorecard database.I didnt intend to pick business schools, but it kind of works out perfectly in regards to selecting for people who are likely not represented proportionally in the college scorecard data.However the very weird world of finacial salaries makes the data far from fullproof, but I would imagine more accurate data would only strengthen the difference between the earnings of the average graduate from one of these programs and the earnings of the average graduate who recieved some form of finacial aid. Particurly the fact that almost all of the Buisness school data was based off starting salaries reported by students, while Gov Scorecard is likely based of tax returns at the end of the first year of work,and includes any potential bonus's/raises recieved throughout the year. 