---
title: "college_final_project"
format: html
execute:
  http: true
---

```{r}
# install.packages("httr2")
library(httr2)
library(tidyverse)
library(plotly)
```

```{r}
school_list <- list()

for (k in 1:500) {
college_request <- request("https://api.data.gov") %>% 
  req_url_path_append("ed") %>% 
  req_url_path_append("collegescorecard") %>% 
  req_url_path_append("v1") %>% 
  req_url_path_append("schools") %>% 
  req_url_query(
    api_key = "sXz5twfrusSSiWtD5EQ35d4e3XNzfiw3m4yHiMuv",
    fields = "id,school.name,latest.student.size,school.state,latest.cost.tuition.in_state,latest.cost.tuition.out_of_state,latest.admissions.admission_rate.overall,latest.cost.attendance.academic_year,latest.completion.completion_rate_4yr_150nt,latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid,latest.earnings.1_yr_after_completion.median,latest.earnings.5_yrs_after_completion.median" ,
    per_page = 100,
    page = k,
  )


  college_response <- req_perform(college_request)
  data <- college_response %>% resp_body_json(simplifyVector = TRUE)
  results <- data$results
  school_list[[k]] <- results
  
  if (nrow(results) < 100) {
    break
  }
}
# install.packages("janitor")

# college_tbl <- college_response %>%
  # resp_body_json(simplifyVector = TRUE) %>%
  # pluck("results") %>%
  # as_tibble()

# college_tbl

colleges <- bind_rows(school_list)

```

```{r}
# colnames(colleges)
```

```{r}

# not sure how accurate the apps & accept columns
# could use this set too it has more rows:
# https://www.kaggle.com/datasets/thedevastator/u-s-department-of-education-college-scorecard-da
#college2 <- read_csv("College_Data.csv")
#college2

#college2 <- college2 %>%
# rename(name = ...1)

#colleges <- colleges %>%
#  rename(name = school.name)
```

```{r}
# merged_colleges <- left_join(college2, colleges, by = "name")

# some duplicates in here
```

```{r}
alt_colleges <- read_csv("alt_college_data.csv")
```

```{r}
alt_colleges <- alt_colleges %>%
  rename(name = INSTNM)

alt_colleges <- alt_colleges %>%
  rename(id = UNITID)
```

```{r}
# merged_colleges_big <- left_join(colleges, alt_colleges, by = c("school.name" = "name"))
# merged_colleges_big <- left_join(colleges, alt_colleges, by = "id")
```

```{r}
# merged_colleges %>%
#  ggplot(aes(x = Private, y = latest.cost.tuition.out_of_state)) +
#    geom_boxplot() +
#    coord_flip() +
#    theme_minimal() +
#    labs(
#      y = "Tuition (Out of State)",
#      x = "Private",
#      title = "Cost of Public vs. Private Schools"
#    )
```

# Part 3: ROI

```{r}
colleges <- colleges %>%
  rename(name = school.name) %>%
  rename(state = school.state) %>%
  rename(five_yr_earnings = latest.earnings.5_yrs_after_completion.median) %>%
  rename(one_yr_earnings = latest.earnings.1_yr_after_completion.median) %>%
  rename(size = latest.student.size) %>%
  rename(instate_tuition = latest.cost.tuition.in_state) %>%
  rename(outstate_tuition = latest.cost.tuition.out_of_state) %>%
  rename(acceptance = latest.admissions.admission_rate.overall) %>%
  rename(avg_cost = latest.cost.attendance.academic_year) %>%
  rename(graduation = latest.completion.completion_rate_4yr_150nt) %>%
  rename(repaid_debt = latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid)
```

## Variables in question

`name`: the institution name
`state`: the institution's state
`instate_tuition`: cost of instate tuition, before aid
`outstate_tuition`: cost of out-of-state tuition, before aid
`acceptance`: overall acceptance rate
`graduation`: overall graduation rate, in four years
`repaid_debt`: proportion of graduates who have fully repaid debt after 2 years
`size:` the most recent number of undergraduate students
`avg_cost:` average cost of attendance, before aid
`one_yr_earnings:` median earnings of graduates after 1 year from graduation
`five_yr_earnings:` median earnings of graduates after 5 years from graduation

## Charts

```{r}
roi_plot1 <- colleges %>%
  ggplot(aes(x = avg_cost, y = graduation, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Graduation Rate",
      color = "Acceptance Rate",
      title = "Graduation Rates vs. Cost of Attendance"
    )
ggplotly(roi_plot1, tooltip = "text")
```

### Observation: there is a clear positive trend between the two variables. This suggests that more expensive colleges might be worth it because there is a higher chance of graduating with a degree.

```{r}
roi_plot2 <- colleges %>%
  ggplot(aes(x = avg_cost, y = one_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 1 Year",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 1 Year"
    )
ggplotly(roi_plot2, tooltip = "text")
```

### Observation: there is a slight positive trend, but it is almost flat. This suggests that a more expensive college might not be worth it in the short-term, using 1-year median earnings as a marker of ROI.

```{r}
roi_plot3 <- colleges %>%
  ggplot(aes(x = avg_cost, y = five_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 5 Years",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 5 Years"
    )
ggplotly(roi_plot3, tooltip = "text")
```

### Observation: the trend is more significant after 5 years. This gives evidence that a more expensive college might be worth it in the long-term, as they can lead to higher-paying careers. 

```{r}
roi_plot4 <- colleges %>%
  ggplot(aes(x = avg_cost, y = acceptance)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Acceptance Rate",
      title = "Acceptance Rate vs. Cost of Attendance",
    )
roi_plot4
```






### Observation: there is a strong negative association between these variables. More expensive colleges are often more selective, with lower acceptance rates. This could mean that the cost of attendance doesn't necessarily predict outcomes, but the prestige of the institution.






```{r}
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)

#Variables in question:
#Major/Field of study = Program title
#Median earnings of pell grant reciepents 4 years after graduating = earnings_4_yr_median_pell
#Median earnings of non pell grant reciepents 4 years after graduating = earnings_4_yr_median_nonpell

url <- "https://api.data.gov/ed/collegescorecard/v1/schools?api_key=sydxfG3MlGadt1EOtAGWOSuSMnYBVFFLvkBeAFiE&latest.programs.cip_4_digit.credential.level=3&fields=id,school.name,latest.programs.cip_4_digit.code,latest.programs.cip_4_digit.title,latest.programs.cip_4_digit.credential.level,latest.programs.cip_4_digit.earnings.4_yr.overall_median_earnings,latest.programs.cip_4_digit.earnings.4_yr.pell_median_earnings,latest.programs.cip_4_digit.earnings.4_yr.nonpell_median_earningsla,6_yrs_after_entry.median_earnings_lowest_tercile"

response <- GET(url)

data <- fromJSON(content(response, "text", encoding = "UTF-8"), flatten = TRUE)

ca_schools2 <- data$results

ca_flat2 <- ca_schools2 %>%
  unnest(latest.programs.cip_4_digit)


ca_clean2 <- ca_flat2 %>%
  rename(
    school_id = id,
    school_name = school.name,
    cip_code = code,
    program_title = title,
    degree_level = credential.level,
    earnings_4_yr_median = earnings.4_yr.overall_median_earnings,
    earnings_4_yr_median_pell = earnings.4_yr.pell_median_earnings,
    earnings_4_yr_median_nonpell = earnings.4_yr.nonpell_median_earnings,
  )

write.csv(ca_clean2, "school_complete_earnings.csv", row.names = FALSE, na = "")

```

```{r}
majors<- ca_clean2%>%
group_by(program_title)%>%
summarize(
majorearningspell = mean(earnings_4_yr_median_pell,na.rm = TRUE), majorearningsnonpell =mean(earnings_4_yr_median_nonpell,na.rm = TRUE),majorearningsoverall =mean(earnings_4_yr_median,na.rm = TRUE),.grops="drop")%>%
mutate(privelege_tax = (majorearningsnonpell - majorearningspell)
  ,privelege_percentage = (privelege_tax/majorearningsoverall))
  top_10_tax <-majors %>%
  slice_max(order_by = privelege_tax, n = 10)
  not_10_tax <-majors %>%
  slice_min(order_by = privelege_tax, n = 10)
ggplot(top_10_tax, aes(x = reorder(program_title, privelege_tax), y = privelege_tax)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Majors by Earnings 'Privilege Tax'",
    subtitle = "Amount less earned by Pell Grant Recipients",
    x = "Major",
    y = "Earnings Gap ($)"
  )
ggplot(not_10_tax, aes(x = reorder(program_title, privelege_tax), y = privelege_tax)) +
  geom_col(fill = "firebrick") + 
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Majors by Earnings 'Privilege Tax'",
    subtitle = "Amount less earned by Pell Grant Recipients",
    x = "Major",
    y = "Earnings Gap ($)"
  )

ggplot(top_10_tax, aes(x = reorder(program_title, privelege_percentage), y = privelege_percentage*100)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Majors by Earnings 'Privilege Tax'",
    subtitle = "Percent less earned by Pell Grant Recipients",
    x = "Major",
    y = "Earnings Gap (%)"
  )



url <- "https://api.data.gov/ed/collegescorecard/v1/schools?api_key=sydxfG3MlGadt1EOtAGWOSuSMnYBVFFLvkBeAFiE&fields=id,school.name,latest.9_yrs_after_entry.median_earnings_lowest_tercile&per_page=100"



response <- GET(url)

data2 <- fromJSON(content(response, "text", encoding = "UTF-8"), flatten = TRUE)

insta <- data2$results



insta_clean2 <- insta %>%
  rename(
    school_id = id,
    school_name = school.name,
    med_family_income = latest.9_yrs_after_entry.median_earnings_lowest_tercile
  )

write.csv(insta_clean2, "school_complete_earnings.csv", row.names = FALSE, na = "")
```



```
Obeservation:The gaps in career earnings are very signifcant but its hard to detect a major trend in the types of majors with the highest gaps. Maybe some data quality issues or something cause I remember the list being different and the data table shows higher median earnings then both pell and non pell so look into that(note)