---
title: "college_final_project"
format: html
execute:
  http: true
  echo: false
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
# install.packages("httr2")
library(httr2)
library(tidyverse)
library(plotly)
library(maps)
```

```{r}
consistent_theme <-theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 15),
    legend.title= element_text(size=15),
    legend.text = element_text(size = 14), 
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
    )
```



```{r}
school_list <- list()

for (k in 1:500) {
college_request <- request("https://api.data.gov") %>% 
  req_url_path_append("ed") %>% 
  req_url_path_append("collegescorecard") %>% 
  req_url_path_append("v1") %>% 
  req_url_path_append("schools") %>% 
  req_url_query(
    api_key = "sXz5twfrusSSiWtD5EQ35d4e3XNzfiw3m4yHiMuv",
    fields = paste(
      "id",
      "school.name",
      "school.ownership",
      "latest.student.size",
      "school.state",
      "latest.cost.tuition.in_state",
      "latest.cost.tuition.out_of_state",
      "latest.admissions.admission_rate.overall",
      "latest.cost.attendance.academic_year",
      "latest.completion.completion_rate_4yr_150nt",
      "latest.earnings.5_yrs_after_completion.median",
      "latest.earnings.1_yr_after_completion.median",
      "latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid",
  sep = ","),
    per_page = 100,
    page = k,
  )


  college_response <- req_perform(college_request)
  data <- college_response %>% resp_body_json(simplifyVector = TRUE)
  results <- data$results
  school_list[[k]] <- results
  
  if (nrow(results) < 100) {
    break
  }
}
colleges <- bind_rows(school_list)

```

# Part 1 How Much More Expensive Are Expensive Colleges?

Before evaluating whether expensive colleges are worth the cost, it is important to understand how college costs vary across institutions. We first examine tuition differences by institution type, the overall distribution of tuition, and geographic patterns across the United States.

## What does "expensive" actually mean?
```{r}
college_cost <- colleges %>%
  mutate(
    school.ownership = factor(
      school.ownership,
      levels = c(1, 2, 3),
      labels = c(
        "Public",
        "Private nonprofit",
        "Private for-profit"
      )
    )
  )

boxpolot_cost <-
  ggplot(
  college_cost,
  aes(
    x = school.ownership,
    y = latest.cost.tuition.out_of_state,
    text = school.name
  )
) +
  geom_boxplot() +
  labs(
    title = "Out-of-State Tuition by Institution Type",
    x = "Institution Type",
    y = "Annual Tuition and Fees ($)" )+
  theme_minimal()

ggplotly(boxpolot_cost, tooltip = "text")
```

## Where are more expensive?
```{r}
state_cost <- colleges %>%
  group_by(school.state) %>%
  summarise(avg_tuition = mean(latest.cost.tuition.out_of_state, na.rm = TRUE)) %>%
  mutate(region = tolower(state.name[match(school.state, state.abb)]))
us_map <- map_data("state")

map_df <- us_map %>%
  left_join(state_cost, by = "region")
```

```{r}
cost_map <- ggplot(map_df,
            aes(
              long,
              lat,
              group = group,
              fill = avg_tuition,
              text = paste(
                "State:",
                tools::toTitleCase(region),
                "<br>Average Tuition: $",
                round(avg_tuition)
              )
            )) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Average Out-of-State Tuition by State", fill = "Tuition ($)") +
  theme_void()
ggplotly(cost_map, tooltip = "text")
```
## What does "expensive" actually mean?
```{r}

median_tuition <- median(

  colleges$latest.cost.tuition.out_of_state,

  na.rm = TRUE

)
ggplot(
  colleges,
  aes(x = latest.cost.tuition.out_of_state)
) +
  geom_histogram(
    bins = 30,
    fill = "#4C78A8",
    color = "white"
  ) +

  geom_vline(

    xintercept = median_tuition,

    color = "red",

    linewidth = 1)+
  labs(
    title = "Distribution of Out-of-State Tuition",
    x = "Out-of-State Tuition ($)",
    y = "Number of Institutions"
  ) +
  theme_minimal()

```


```{r}
colleges <- colleges %>%
  rename(name = school.name) %>%
  rename(state = school.state) %>%
  rename(five_yr_earnings = latest.earnings.5_yrs_after_completion.median) %>%
  rename(one_yr_earnings = latest.earnings.1_yr_after_completion.median) %>%
  rename(size = latest.student.size) %>%
  rename(instate_tuition = latest.cost.tuition.in_state) %>%
  rename(outstate_tuition = latest.cost.tuition.out_of_state) %>%
  rename(acceptance = latest.admissions.admission_rate.overall) %>%
  rename(avg_cost = latest.cost.attendance.academic_year) %>%
  rename(graduation = latest.completion.completion_rate_4yr_150nt) %>%
  rename(repaid_debt = latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid)
```

## Variables in question

`name`: the institution name \n `state`: the institution's state \n `instate_tuition`: cost of instate tuition, before aid \n `outstate_tuition`: cost of out-of-state tuition, before aid \n `acceptance`: overall acceptance rate \n `graduation`: overall graduation rate, in four years \n `repaid_debt`: proportion of graduates who have fully repaid debt after 2 years \n `size:` the most recent number of undergraduate students \n `avg_cost:` average cost of attendance, before aid \n `one_yr_earnings:` median earnings of graduates after 1 year from graduation \n `five_yr_earnings:` median earnings of graduates after 5 years from graduation



# Part 2: Do expensive colleges produce better outcomes? 

In this part, we investigate whether more expensive colleges are worth the cost by producing better outcomes, that is, leading to higher graduation rate and higher income after graduation. For income after graduation, we examined the median income one years after graduation and five years after graduation to see whether the trend persists over time.  

## Relationship between tuition and graduation rate 

```{r}

# Graduation rate VS In state tuition 

grad_rate_inStateTuition<-colleges%>%
  ggplot(aes(x=instate_tuition,
             y=graduation)
         )+
  geom_point(aes(text = name))+
  geom_smooth(method = "lm")+
  labs(x="In-state Tuition",
       y="Graduation Rate",
       title="Graduation Rate VS In-state Tuition")+
   consistent_theme
 


# Graduation rate VS Out of state tuition 

grad_rate_outStateTuition<-colleges%>%
  ggplot(aes(x=outstate_tuition,
             y=graduation)
         )+
  geom_point(aes(text = name))+
  geom_smooth(method = "lm")+
  labs(x="Out-of-state Tuition",
       y="Graduation Rate",
       title="Graduation Rate VS Out-of-state Tuition")+
  consistent_theme
  


subplot(
  ggplotly(grad_rate_inStateTuition, tooltip = "text"),
  ggplotly(grad_rate_outStateTuition, tooltip = "text"),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE
) %>%
  layout(
    title = list(
      text = "<b>Graduation Rate vs Tuition</b>",
      font = list(size = 24),
      x = 0.5
    )
  )

```

Here, we observed a positive correlation between graduation rate and tuition (both in-state and out-of-state). This means that the higher the tuition, the higher the graduation rate. This is probably because institutions with higher tuition fee are usually better funded and thus can offer more resources to help students earn their degree. Another reason may be that students who can afford attending colleges with higher tuition usually come from a more well-off socioeconomic background and thus can get more support and resources from their family in finishing college. Moreover, students who pay more to attend colleges would feel more pressured to finish their degrees with the mind that "I can't waste this money".  


## Relationship between tuition and median earnings 1 year after graduation 

```{r}

# One year earnings VS In state tuition

oneYr_earn_inStateTuition<-colleges%>%
  ggplot(aes(x=instate_tuition,
             y=one_yr_earnings)
         )+
  geom_point(aes(text = name))+
  geom_smooth(method = "lm")+
  labs(x="In-state Tuition",
       y="Median earnings 1 year after graduation",
      title= "Median After-One-year Earnings VS \n In-state Tuition"
       )+
   consistent_theme
 
  

# One year earnings VS Out of state tuition 

oneYr_earn_outStateTuition<-colleges%>%
  ggplot(aes(x=outstate_tuition,
             y=one_yr_earnings)
         )+
  geom_point(aes(text = name))+
  geom_smooth(method = "lm")+
  labs(x="Out-of-state Tuition",
       y="Median earnings 1 year after graduation",
       title= "Median After-One-year Earnings VS \n Out-of-state Tuition"
       )+
  consistent_theme
  


# 1 year earnings VS tuition
subplot(
  ggplotly(oneYr_earn_inStateTuition, tooltip = "text"),
  ggplotly(oneYr_earn_outStateTuition, tooltip = "text"),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE
) %>%
  layout(
    title = list(
      text = "<b> Median After-One-year Earnings VS \n Tuition </b>",
      font = list(size = 24),
      x = 0.5
    )
  )

```

From the graph, there is a median strong positive relationship between the median earnings 1 year after graduation of the graduates and both the in-state and out-of-state tuition, meaning that higher the tuition, higher the income soon after graduation. One's first year of earnings after graduation usually comes from their first job. Thus, our results show that those who attend colleges with higher tuition fee will be able to find a first job with higher earnings, indicating that the outcome is worth the cost. 


```{r}


# Five year earning VS In state tuition

fiveYr_earn_inStateTuition<-colleges%>%
  ggplot(aes(x=instate_tuition,
             y=five_yr_earnings)
         )+
  geom_point(aes(text = name))+
  geom_smooth(method = "lm")+
  labs(x="In-state Tuition",
       y="Median earnings 5 years after graduation",
       title="Median After-Five-years Earnings VS \n In-state Tuition")+
  consistent_theme



# Five year earnings VS Out of State tuition 

fiveYr_earn_outStateTuition<-colleges%>%
  ggplot(aes(x=outstate_tuition,
             y=five_yr_earnings)
         )+
  geom_point(aes(text = name))+
  geom_smooth(method = "lm")+
  labs(x="Out-of-state Tuition",
       y="Median earnings 5 year after graduation",
       title="Median After-Five-years Earnings VS \n Out-of-state Tuition")+
  consistent_theme
  



subplot(
  ggplotly(fiveYr_earn_inStateTuition, tooltip = "text"),
  ggplotly(fiveYr_earn_outStateTuition, tooltip = "text"),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE
) %>%
  layout(
    title = list(
      text = "<b> Median After-Five-year Earnings VS \n Tuition</b>",
      font = list(size = 24),
      x = 0.5
    )
  )



```

Our graph shows that there is a even stronger positive relationship between tuition (both in-state and out-of-state) and median earnings five years after graduation compared to one year after graduation. This indicates that the trend for higher tuition higher outcome persists over time and gets even more obvious. This is probably either because as years go by, one's earnings will increase with things like promotion, or the person is able to switch to a job with higher earnings. Thus, attending college with high tuition fee will not only help graduates secure a first job with better earnings but also help people stay in this high-salary job and earn more and more over time or find another job with higher salary. 



# Part 3: ROI


## Charts

```{r}
roi_plot1 <- colleges %>%
  ggplot(aes(x = avg_cost, y = graduation, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Graduation Rate",
      color = "Acceptance Rate",
      title = "Graduation Rates vs. Cost of Attendance"
    )
ggplotly(roi_plot1, tooltip = "text")
```

Observation: there is a clear positive trend between the two variables. This suggests that more expensive colleges might be worth it because there is a higher chance of graduating with a degree.

```{r}
roi_plot2 <- colleges %>%
  ggplot(aes(x = avg_cost, y = one_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 1 Year",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 1 Year"
    )
ggplotly(roi_plot2, tooltip = "text")
```

Observation: there is a slight positive trend, but it is almost flat. This suggests that a more expensive college might not be worth it in the short-term, using 1-year median earnings as a marker of ROI.

```{r}
roi_plot3 <- colleges %>%
  ggplot(aes(x = avg_cost, y = five_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 5 Years",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 5 Years"
    )
ggplotly(roi_plot3, tooltip = "text")
```

Observation: the trend is more significant after 5 years. This gives evidence that a more expensive college might be worth it in the long-term, as they can lead to higher-paying careers.

```{r}
colleges <- colleges %>%
  mutate(
    repaid_debt = parse_number(repaid_debt)
  )
roi_plot4 <- colleges %>%
  ggplot(aes(x = avg_cost, y = repaid_debt, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Proportion of Graduates with Repaid Debt",
      color = "Acceptance Rate",
      title = "Repaid Debt vs. Cost of Attendance",
    )
ggplotly(roi_plot4, tooltip = "text")
```

Observation: there is a trend of higher-costing colleges yielding better results for debt repayment. However, this is influenced by some of the outliers on the top right, which blurs the results.

```{r}
roi_plot5 <- colleges %>%
  ggplot(aes(x = avg_cost, y = acceptance)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Acceptance Rate",
      title = "Acceptance Rate vs. Cost of Attendance",
    )
roi_plot5
```

Observation: there is a strong negative association between these variables. More expensive colleges are often more selective, with lower acceptance rates. This could mean that the cost of attendance doesn't necessarily predict outcomes, but the prestige of the institution.

## Looking at top 500 college data

```{r}
library(stringr)
top_colleges <- read_csv("clean_top_colleges.csv")
```
  
```{r}
ranked_colleges <- top_colleges %>%
  left_join(colleges, by = "name")

ranked_colleges <- ranked_colleges %>%
  select(c("name", "rank", "state", "acceptance", "graduation", "avg_cost", "one_yr_earnings", "five_yr_earnings"))
```

The new data set, `ranked_colleges`, contains the information from the College Scorecard API for just the top 500 colleges, according to this ranking of colleges from 2022. 

### Charts

```{r}
roi_pt2_plot1 <- ranked_colleges %>%
  ggplot(aes(x = avg_cost, y = one_yr_earnings, color = rank)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 1 Year",
      color = "Rank",
      title = "Median Earnings After 1 Year vs. Cost of Attendance",
      subtitle = "For Top 500 Schools"
    )
ggplotly(roi_pt2_plot1, tooltip = "text")
```

Observation: there is not a clear trend here, though the line of best fit is slightly positive. There are also two clear clusters here, likely public schools on the left (cheaper) and private schools on the right (more expensive). 

```{r}
roi_pt2_plot2 <- ranked_colleges %>%
  ggplot(aes(x = avg_cost, y = five_yr_earnings, color = rank)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 5 Years",
      color = "Rank",
      title = "Median Earnings After 5 Years vs. Cost of Attendance",
      subtitle = "For Top 500 Schools"
    )
ggplotly(roi_pt2_plot2, tooltip = "text")
```

Observation: after 5 years, there is a slightly steeper trend line, though it's not clear. This is likely influenced by outliers on the upper right hand side of the graph. Other than those schools, the trend is basically flat, suggesting less expensive schools might be worth it because the outcomes are similar for schools across cost.

Summary: unless you attend one of the outlier schools (MIT, Stanford, Harvey Mudd, Cal Tech), it doesn't really make a difference if you pay more. The less expensive schools probably have a better return on investment, but this perspective does not take into account financial aid.

<<<<<<< HEAD

