---
title: "college_final_project"
format: html
execute:
  http: true
  echo: false
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
# install.packages("httr2")
library(httr2)
library(tidyverse)
library(plotly)
library(maps)
```

```{r}
school_list <- list()

for (k in 1:500) {
college_request <- request("https://api.data.gov") %>% 
  req_url_path_append("ed") %>% 
  req_url_path_append("collegescorecard") %>% 
  req_url_path_append("v1") %>% 
  req_url_path_append("schools") %>% 
  req_url_query(
    api_key = "sXz5twfrusSSiWtD5EQ35d4e3XNzfiw3m4yHiMuv",
    fields = paste(
      "id",
      "school.name",
      "school.ownership",
      "latest.student.size",
      "school.state",
      "latest.cost.tuition.in_state",
      "latest.cost.tuition.out_of_state",
      "latest.admissions.admission_rate.overall",
      "latest.cost.attendance.academic_year",
      "latest.completion.completion_rate_4yr_150nt",
      "latest.earnings.5_yrs_after_completion.median",
      "latest.earnings.1_yr_after_completion.median",
      "latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid",
  sep = ","),
    per_page = 100,
    page = k,
  )


  college_response <- req_perform(college_request)
  data <- college_response %>% resp_body_json(simplifyVector = TRUE)
  results <- data$results
  school_list[[k]] <- results
  
  if (nrow(results) < 100) {
    break
  }
}
colleges <- bind_rows(school_list)

```

```{r}
#glimpse(colleges)
```

```{r}

# not sure how accurate the apps & accept columns
# could use this set too it has more rows:
# https://www.kaggle.com/datasets/thedevastator/u-s-department-of-education-college-scorecard-da
#college2 <- read_csv("College_Data.csv")
#college2

#college2 <- college2 %>%
# rename(name = ...1)

#colleges <- colleges %>%
#  rename(name = school.name)
```

```{r}
# merged_colleges <- left_join(college2, colleges, by = "name")

# some duplicates in here
```

```{r}
alt_colleges <- read_csv("alt_college_data.csv")
```

```{r}
alt_colleges <- alt_colleges %>%
  rename(name = INSTNM)

alt_colleges <- alt_colleges %>%
  rename(id = UNITID)
```

```{r}
# merged_colleges_big <- left_join(colleges, alt_colleges, by = c("school.name" = "name"))
# merged_colleges_big <- left_join(colleges, alt_colleges, by = "id")
```

```{r}
# merged_colleges %>%
#  ggplot(aes(x = Private, y = latest.cost.tuition.out_of_state)) +
#    geom_boxplot() +
#    coord_flip() +
#    theme_minimal() +
#    labs(
#      y = "Tuition (Out of State)",
#      x = "Private",
#      title = "Cost of Public vs. Private Schools"
#    )
```

```{r}
#colnames(colleges)
```

# Part 1 How Much More Expensive Are Expensive Colleges?

Before evaluating whether expensive colleges are worth the cost, it is important to understand how college costs vary across institutions. We first examine tuition differences by institution type, the overall distribution of tuition, and geographic patterns across the United States.

## What does "expensive" actually mean?
```{r}
college_cost <- colleges %>%
  mutate(
    school.ownership = factor(
      school.ownership,
      levels = c(1, 2, 3),
      labels = c(
        "Public",
        "Private nonprofit",
        "Private for-profit"
      )
    )
  )

boxpolot_cost <-
  ggplot(
  college_cost,
  aes(
    x = school.ownership,
    y = latest.cost.tuition.out_of_state,
    text = school.name
  )
) +
  geom_boxplot() +
  labs(
    title = "Out-of-State Tuition by Institution Type",
    x = "Institution Type",
    y = "Annual Tuition and Fees ($)" )+
  theme_minimal()

ggplotly(boxpolot_cost, tooltip = "text")
```

## Where are more expensive?
```{r}
state_cost <- colleges %>%
  group_by(school.state) %>%
  summarise(avg_tuition = mean(latest.cost.tuition.out_of_state, na.rm = TRUE)) %>%
  mutate(region = tolower(state.name[match(school.state, state.abb)]))
us_map <- map_data("state")

map_df <- us_map %>%
  left_join(state_cost, by = "region")
```

```{r}
cost_map <- ggplot(map_df,
            aes(
              long,
              lat,
              group = group,
              fill = avg_tuition,
              text = paste(
                "State:",
                tools::toTitleCase(region),
                "<br>Average Tuition: $",
                round(avg_tuition)
              )
            )) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Average Out-of-State Tuition by State", fill = "Tuition ($)") +
  theme_void()
ggplotly(cost_map, tooltip = "text")
```
## What does "expensive" actually mean?
```{r}

median_tuition <- median(

  colleges$latest.cost.tuition.out_of_state,

  na.rm = TRUE

)
ggplot(
  colleges,
  aes(x = latest.cost.tuition.out_of_state)
) +
  geom_histogram(
    bins = 30,
    fill = "#4C78A8",
    color = "white"
  ) +

  geom_vline(

    xintercept = median_tuition,

    color = "red",

    linewidth = 1)+
  labs(
    title = "Distribution of Out-of-State Tuition",
    x = "Out-of-State Tuition ($)",
    y = "Number of Institutions"
  ) +
  theme_minimal()

```


#Part 2: Do expensive colleges produce better outcomes? 


## Relationship between tuition and earnings

```{r}

library(ggplot2)

 consistent_theme <-theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 15),
    legend.title= element_text(size=15),
    legend.text = element_text(size = 14), 
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
    )
 

# One year earnings VS In state tuition

oneYr_earn_inStateTuition<-colleges%>%
  ggplot(aes(x=latest.cost.tuition.in_state,
             y=latest.earnings.1_yr_after_completion.median)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="In-state Tuition",
       y="Median earnings 1 year after graduation",
       title="Median After-One-year Earnings VS \n In-state Tuition")+
   consistent_theme
 
  
oneYr_earn_inStateTuition


# One year earnings VS Out of state tuition 

oneYr_earn_outStateTuition<-colleges%>%
  ggplot(aes(x=latest.cost.tuition.out_of_state,
             y=latest.earnings.1_yr_after_completion.median)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="Out-of-state Tuition",
       y="Median earnings 1 year after graduation",
       title="Median After-One-year Earnings VS \n Out-of-state Tuition")+
  consistent_theme
  
oneYr_earn_outStateTuition


# Five year earning VS In state tuition

fiveYr_earn_inStateTuition<-colleges%>%
  ggplot(aes(x=latest.cost.tuition.in_state,
             y=latest.earnings.5_yrs_after_completion.median)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="In-state Tuition",
       y="Median earnings 5 years after graduation",
       title="Median After-Five-years Earnings VS \n In-state Tuition ")+
  consistent_theme
  
fiveYr_earn_inStateTuition



# Five year earnings VS Out of State tuition 

fiveYr_earn_outStateTuition<-colleges%>%
  ggplot(aes(x=latest.cost.tuition.out_of_state,
             y=latest.earnings.5_yrs_after_completion.median)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="Out-of-state Tuition",
       y="Median earnings 5 year after graduation",
       title="Median After-Five-years Earnings VS \n Out-of-state Tuition")+
  consistent_theme
  
fiveYr_earn_outStateTuition


```


## Relationship between tuition and graduation rate 

```{r}

# Graduation rate VS In state tuition 

grad_rate_inStateTuition<-colleges%>%
  ggplot(aes(x=latest.cost.tuition.in_state,
             y=latest.completion.completion_rate_4yr_150nt)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="In-state Tuition",
       y="Graduation Rate",
       title="Graduation Rate VS In-state Tuition")+
   consistent_theme
 
  
grad_rate_inStateTuition


# Graduation rate VS Out of state tuition 

grad_rate_outStateTuition<-colleges%>%
  ggplot(aes(x=latest.cost.tuition.out_of_state,
             y=latest.completion.completion_rate_4yr_150nt)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="Out-of-state Tuition",
       y="Graduation Rate",
       title="Graduation Rate VS Out-of-state Tuition")+
  consistent_theme
  
grad_rate_outStateTuition


```



## Relationship between cost of attendance and graduation rate

```{r}


grad_rate_costAttendance<-colleges%>%
  ggplot(aes(x=latest.cost.attendance.academic_year,
             y=latest.completion.completion_rate_4yr_150nt)
         )+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x="Cost of attendance",
       y="Graduation Rate",
       title="Graduation Rate VS Cost of Attendance")+
   consistent_theme
 
  
grad_rate_costAttendance


```











# Part 3: ROI

```{r}
colleges <- colleges %>%
  rename(name = school.name) %>%
  rename(state = school.state) %>%
  rename(five_yr_earnings = latest.earnings.5_yrs_after_completion.median) %>%
  rename(one_yr_earnings = latest.earnings.1_yr_after_completion.median) %>%
  rename(size = latest.student.size) %>%
  rename(instate_tuition = latest.cost.tuition.in_state) %>%
  rename(outstate_tuition = latest.cost.tuition.out_of_state) %>%
  rename(acceptance = latest.admissions.admission_rate.overall) %>%
  rename(avg_cost = latest.cost.attendance.academic_year) %>%
  rename(graduation = latest.completion.completion_rate_4yr_150nt) %>%
  rename(repaid_debt = latest.repayment.2_yr_bb_fed_repayment.ugcomp.fullypaid)
```

## Variables in question

`name`: the institution name `state`: the institution's state `instate_tuition`: cost of instate tuition, before aid `outstate_tuition`: cost of out-of-state tuition, before aid `acceptance`: overall acceptance rate `graduation`: overall graduation rate, in four years `repaid_debt`: proportion of graduates who have fully repaid debt after 2 years `size:` the most recent number of undergraduate students `avg_cost:` average cost of attendance, before aid `one_yr_earnings:` median earnings of graduates after 1 year from graduation `five_yr_earnings:` median earnings of graduates after 5 years from graduation

## Charts

```{r}
roi_plot1 <- colleges %>%
  ggplot(aes(x = avg_cost, y = graduation, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Graduation Rate",
      color = "Acceptance Rate",
      title = "Graduation Rates vs. Cost of Attendance"
    )
ggplotly(roi_plot1, tooltip = "text")
```

### Observation: there is a clear positive trend between the two variables. This suggests that more expensive colleges might be worth it because there is a higher chance of graduating with a degree.

```{r}
roi_plot2 <- colleges %>%
  ggplot(aes(x = avg_cost, y = one_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 1 Year",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 1 Year"
    )
ggplotly(roi_plot2, tooltip = "text")
```

### Observation: there is a slight positive trend, but it is almost flat. This suggests that a more expensive college might not be worth it in the short-term, using 1-year median earnings as a marker of ROI.

```{r}
roi_plot3 <- colleges %>%
  ggplot(aes(x = avg_cost, y = five_yr_earnings, color = acceptance)) +
    geom_point(aes(text = name)) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Median Earnings After 5 Years",
      color = "Acceptance Rate",
      title = "Median Earnings vs. Cost of Attendance",
      subtitle = "After 5 Years"
    )
ggplotly(roi_plot3, tooltip = "text")
```

### Observation: the trend is more significant after 5 years. This gives evidence that a more expensive college might be worth it in the long-term, as they can lead to higher-paying careers.

```{r}
roi_plot4 <- colleges %>%
  ggplot(aes(x = avg_cost, y = acceptance)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    theme_minimal() +
    labs(
      x = "Average Cost of Attendance",
      y = "Acceptance Rate",
      title = "Acceptance Rate vs. Cost of Attendance",
    )
roi_plot4
```

### Observation: there is a strong negative association between these variables. More expensive colleges are often more selective, with lower acceptance rates. This could mean that the cost of attendance doesn't necessarily predict outcomes, but the prestige of the institution.
<<<<<<< HEAD



